---
title: "GO term enrichment analysis"
author: "ans Carlo von Vegesack"
date: "2024-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setup for first installation
Only execute if you re-installed R or never installed BiocManager
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
BiocManager::install("pheatmap")
BiocManager::install("topGO")
```

### Normal start
Set the working directory as you need it. In my case I set the working directory to a folder called "final_project". Inside this folder is a folder called "count_matrix". "count_matrix" has the ReadsPerGene files. You can see this also in the Load Count Matrix part.
```{r}
# Load necessary libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(topGO))
suppressPackageStartupMessages(library(dplyr))

# Set the working directory
setwd("~/uni/autumn-24/final_project/")
```
## Convert the mapping file
```{r}
# Load necessary libraries
library(dplyr)

# Load the GAF file
gaf_file <- "data/sgd.gaf"  # Adjust the path if needed
gaf_data <- read.delim(gaf_file, comment.char = "!", header = FALSE, sep = "\t", stringsAsFactors = FALSE)

# Extract relevant columns: 11 = Systematic Name, 5 = GO Term
gene_to_go <- gaf_data[, c(11, 5)]
colnames(gene_to_go) <- c("Gene", "GO")

# Extract the first identifier before the pipe
gene_to_go$Gene <- sapply(strsplit(gene_to_go$Gene, "\\|"), `[`, 1)



# Ensure there are no extra spaces or invalid characters
gene_to_go$Gene <- trimws(gene_to_go$Gene)  # Trim whitespace
gene_to_go$GO <- trimws(gene_to_go$GO)  # Trim whitespace

# Preview the cleaned data
head(gene_to_go)

cleaned_gene_to_go <- gene_to_go
# Group by Gene and collapse GO terms
geneID2GO <- cleaned_gene_to_go %>%
  group_by(Gene) %>%
  mutate(GO = paste(unique(GO), collapse = ","), .groups = "drop")

geneID2GO <- geneID2GO %>% distinct(Gene, GO)
# Save the collapsed file
write.table(geneID2GO, file = "data/collapsed_geneID2GO.tsv", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)


```

## Load Mapping file and create character vector
```{r}
geneID2GO <- readMappings("data/collapsed_geneID2GO.tsv")

geneUniverse <- names(geneID2GO)
```


## Load Gene Expression Data
```{r}
data <- read_csv("data/yeast_results_acetic_45_200.csv")
```

## Filter for statistically significant upregulated genes
```{r}
up_genes <- data %>% 
  filter(padj <= 0.05 & log2FoldChange >= 0)

# Filter for statistically signficant downregulated genes
down_genes <- data %>% 
  filter(padj <= 0.05 & log2FoldChange <= 0)
```

## Extract Gne ids
```{r}
upregulated_genes <- as.character(up_genes$gene_id)
downregulated_genes <- as.character(down_genes$gene_id)
```

## Get binary values depending on if a gene is upregulated or not (or downregulated or not)
```{r}
up_gene_list <- factor(as.integer(geneUniverse %in% upregulated_genes))
down_gene_list <- factor(as.integer(geneUniverse %in% downregulated_genes))

# Set names for the gene list.
names(up_gene_list) <- geneUniverse
names(down_gene_list) <- geneUniverse
```

```{r}
# Check the length of geneUniverse
length(geneUniverse)

# Check the number of matches
sum(geneUniverse %in% upregulated_genes)  # Should be > 0
sum(geneUniverse %in% downregulated_genes)  # Should be > 0

# Check a preview of gene IDs in each list
head(geneUniverse)
head(upregulated_genes)
head(downregulated_genes)

```

# GO Term Enrichment
```{r}
up_GO_data <- new("topGOdata", 
                  description = "yeast_acidic", 
                  ontology = "BP", 
                  allGenes = up_gene_list,
                  annot = annFUN.gene2GO,
                  gene2GO = geneID2GO)

# Build the GOdata object in topGO for downregulated
down_GO_data <- new("topGOdata",
                    description = "yeast_acidic",
                    ontology = "BP",
                    allGenes = down_gene_list,
                    annot = annFUN.gene2GO,
                    gene2GO = geneID2GO)
```
```{r}
# Perform stats for upregulated data
up_result <- runTest(up_GO_data,
                     algorithm = "weight01",
                     statistic = "fisher")

# Perform stats for downregulated data
down_result <- runTest(down_GO_data,
                       algorithm = "weight01",
                       statistic = "fisher")
```
```{r}
# Extract a summary of upregulated results
up_GO <- GenTable(up_GO_data,
                       weight01 = up_result,
                       orderBy = "up_result",
                       ranksOf = "up_result",
                       topNodes = 50)

# Extract a summary of downregulated results
down_GO <- GenTable(down_GO_data,
                         weight01 = down_result,
                         orderBy = "down_result",
                         ranksOf = "down_result",
                         topNodes = 50)
```


# Visualizing TopGO Analyses
```{r}
# Filter out any non-significant data and calculate the gene ratio
down_GO_filtered <- down_GO %>%
  mutate(GeneRatio = Significant/Annotated, weight01 = as.numeric(weight01)) %>%
  filter(weight01 <= 0.05) %>%
  head(n = 20)
```

```{r}
down_GO_filtered %>% 
  ggplot(aes(x = Term, y = GeneRatio)) +
  geom_col(width = 0.05) + 
  geom_point(size = 3) +
  coord_flip() # Flip the axes so the x-axis labels are readable
```
```{r}
# First, let's arrange the data based on the enrichment ratio. 
down_GO_filtered_arranged <- down_GO_filtered %>% 
  arrange(GeneRatio) %>%
  mutate(Term = factor(Term))

# Now let's extract the order of the term column
order_term <- down_GO_filtered_arranged %>% 
  pull(Term) # pull() extracts a column as a vector
```

```{r}
down_GO_filtered_arranged %>% 
  ggplot(aes(x= Term, y = GeneRatio, colour = weight01)) +
  geom_col(width = 0.05) +
  geom_point(size = 3) +
  coord_flip() +
  scale_x_discrete(limits = order_term) + 
  scale_colour_gradient(low = "red", high = "blue")
```
```{r}
# Filter out any non-significant data and calculate the gene ratio
up_GO_filtered <- up_GO %>%
  mutate(GeneRatio = Significant/Annotated, weight01 = as.numeric(weight01)) %>%
  filter(weight01 <= 0.05) %>%
  head(n = 20)
```
```{r}
up_GO_filtered %>% 
  ggplot(aes(x = Term, y = GeneRatio)) +
  geom_col(width = 0.05) + 
  geom_point(size = 3) +
  coord_flip() # Flip the axes so the x-axis labels are readable
```
```{r}
# First, arrange the data based on the enrichment ratio. 
up_GO_filtered_arranged <- up_GO_filtered %>% 
  arrange(GeneRatio) %>%
  mutate(Term = factor(Term))

# extract the order of the term column
order_term_up <- up_GO_filtered_arranged %>% 
  pull(Term) # pull() extracts a column as a vector
```

```{r}

up_GO_filtered_arranged %>% 
  ggplot(aes(x= Term, y = GeneRatio, colour = weight01)) +
  geom_col(width = 0.05) +
  geom_point(aes(size= Significant)) +
  coord_flip() +
  scale_x_discrete(limits = order_term_up) + 
  scale_colour_gradient(low = "red", high = "blue") +
    theme_light() +
  labs(x = "GO Term Description", y = "Enrichment Ratio", color = "P-value", size = "Number of Significant Genes") +
  theme(panel.border = element_rect(color = "black"), panel.grid = element_line(colour = "grey96")) +
   scale_y_continuous(limits = c(0, 1.1), breaks = seq(0, 1, 0.25), expand = c(0, 0))


```
```{r}
# Add labels to upregulated and downregulated dataframes
up_GO <- up_GO %>% 
  mutate(up_down = "UP")

down_GO <- down_GO %>% 
  mutate(up_down = "DOWN")

# Make a joined dataframe
joined_GO_filtered_arranged <- bind_rows(up_GO, down_GO) %>%
  filter(weight01 <= 0.05) %>%
  mutate(GeneRatio = Significant/Annotated, weight01 = as.numeric(weight01)) %>%
  arrange(GeneRatio) %>%
  mutate(Term = factor(Term)) %>%
  head(n = 40)

# Extract the column order
order_term_joined <- joined_GO_filtered_arranged %>% 
  pull(Term)

joined_GO_filtered_arranged %>% 
  ggplot(aes(x= Term, y = GeneRatio, color = weight01)) +
  geom_point(aes(size= Significant)) +
  coord_flip() +
  scale_x_discrete(limits = order_term_joined) +
  scale_color_gradient(low = "red", high = "blue") +
  theme_light() +
  labs(x = "GO Term Description", y = "Enrichment Ratio", color = "P-value", size = "Number of Significant Genes") +
  theme(panel.border = element_rect(color = "black"), 
        panel.grid = element_line(colour = "grey96"), 
        strip.background = element_rect(colour = "black")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2), expand = c(0, 0)) +
  facet_grid(.~ up_down)
```

